The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum

### 2014-03-20 cordillera-climate.tex
Typesetting: fix quotes and preceding space

Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate. However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a~``glacial index'' weighting function derived from ice core \chem{\delta^{18}O} records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.

@@

The Shallow Shelf Approximation (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}. SIA and SSA velocities are computed by finite difference methods on a~10\,km resolution horizontal grid of 300 by 150 points (the modelling domain). Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a~depth of 1\,km. Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model. Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran ice sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations. These new simulations use a~“hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A~single temperature offset of 5\,\unit{{\degree}C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

@@

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$. This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR). These qualitative “best” runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

% fig:biatm
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatm}
	\end{center}
	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row). In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data. Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
	\label{fig:biatm}
\end{figure*}

>>>>

Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate. However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core \chem{\delta^{18}O} records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.

@@

The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}. SIA and SSA velocities are computed by finite difference methods on a~10\,km resolution horizontal grid of 300 by 150 points (the modelling domain). Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a~depth of 1\,km. Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model. Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran ice sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations. These new simulations use a ``hybrid'' climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A~single temperature offset of 5\,\unit{{\degree}C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

@@

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$. This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR). These qualitative ``best'' runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

% fig:biatm
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatm}
	\end{center}
	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using ``hybrid'' climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row). In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data. Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
	\label{fig:biatm}
\end{figure*}
